_____________________________

Topic 1 Practice

_____________________________

Create a separate r chunk for each question

1) type out and run the equation for 100 multiplied by 71

```{r}
100 * 71
```

2) what is 8.8819 multiplied by 9.3147591? Using the round function, round your answer to 3 places after the decimal

```{r}
num <- 8.8819 * 9.3147591
round(num, 3)
```

3) create two objects named class 1 and class 2. Assign class 1 the value of 12 and class 2 the value of 16
```{r}
class_1 <- 12
class_2 <- 16
```


4) add class 1 and class 2 together

```{r}
class_1 + class_2
```


5) We want to track how many students are in 5 sections of History 101. Create a vector with the following information
  a) name the vector "History_Sections", the number of students in each class are as follows c1 = 12, c2=16, c3=14, c4= 20 c5=22

 ```{r}
 History_Sections <- c(12, 16, 14, 20, 22)
 ```

  
  b) Oops! We miscounted the students in class 3. using the element function, change c3 to 18 students

  ```{r}
  History_Sections[3] <- 18
  ```
  

  c) Each professor of History 101 is collecting money for a fieldtrip. They are going to different museums and each museum charges a different fee. Create a vector that shows how much money has to be collected for each class.  The fees are as follows: c1 - $5, c2 = $10, c3 = $5, c4= $12, c5= $18
  ```{r}
  fees <- c(5, 10, 5, 12, 18)
  overall_cost <- History_Sections * fees
  ```



