---
title: "Topic 4 activity"
output: html_notebook
---
1) #We have 20 office supplies in a box. 10 staplers, 5 paperclips, 2 pens, 3 pencils. Create a vector that shows how many of each office supply is in the box

```{r}

office <- rep (c("staplers", "paperclips", "pens", "pencils"), time = c(10, 5, 2, 3))
```

2) show the frequency of each office supply in the box with one function

```{r}
table (office)
```
3) What are the proportions of each office supply?

```{r}
prop.table(table(office))
```
4) Pull a random sample of 4 office supplies from the box. Set seed to 100. What office supplies are pulled from the box?
```{r}
set.seed(100)
sample(office, 4)
```
5) a)Pull another sample of 20 office supplies but replace the office supplies each time. 
   b)create an object for that sample
   c)replicate this event 50 times
   d)set seed to 100. 

```{r}

set.seed(100)
office_event<- replicate(50, sample (office, 20, replace = T))
                          
```

6) Create a table that shows the probability of pulling each supply after 50 trials. Is it similar to our original table? 

```{r}
prop.table(table(office_event))

```

7) We have the following temperature data for 10 days in a given month

```{r}
temp <- c(80, 90, 71, 31, 67, 82, 74, 93, 99, 91)
```

8) Based on this data, and assuming a normal distribution, what is the probability that there's one day that is 73 degrees or lower?

```{r}
mean(temp)
sd(temp)
pnorm (73, mean = mean(temp), sd = sd (temp))
#~40%.
```
9) What is the probability that a day falls between 73 and 83 degrees?
```{r}
pnorm(83, mean = mean(temp), sd = sd(temp))
pnorm(73, mean = mean(temp), sd = sd(temp))
(pnorm(83, mean = mean(temp), sd = sd(temp)) - pnorm(73, mean = mean(temp), sd = sd(temp)))
```
10) What is the probability that there is a day hotter than 83 degrees

```{r}
1-pnorm(83, mean = mean(temp), sd = sd(temp))
```

