---
title: "Topic 3" Plotting with ggplot2 In Class Activity"
output: html_notebook
---

```{r}
library(dplyr)
library(tidyr)
library (ggplot2)
setwd("C:/Users/owner/Desktop/Fall 2024 Binghamton/DIDA 130/DIDA 130/Data")
#change the working directory to where you store your data

#1) Create a data frame for the Global import of cocoa beans


cocoa_import <- read.csv("Global import of cocoa beans.csv")
```

#tell r to read the data file

#2) summarise the data to show the total import in tonnes per year for all regions
#hint--refer to tutorial 2: group by and summarise

```{r}
cocoa_import %>% group_by(Year)

global_cocoa_import <- cocoa_import %>%
  group_by(Year) %>%
  summarise(total_import = sum(Import.in.tonnes))
```

#create a new dataframe that is created from the original data frame, cocoa_import. then you are grouping the cocoa_import data frame by Year, using the group_by function from DPLYR.  Then you are using the summarise function to create a new variable called "total_import" that is created (=) by adding up all of tones per region per year using the sum function. 
  

#3)create the base of a plot graph in ggplot with just the aesthetic layer for x and y

```{r}
ggplot(global_cocoa_import, aes(x=Year, y=total_import))
```

#4) create a full plot graph in ggplot with blue points

```{r}
ggplot(global_cocoa_import, aes(x = Year, y = total_import)) +
  geom_point(color = "blue")  
```


#5) add a line to your point graph to a line graph

```{r}
ggplot(global_cocoa_import, aes(x = Year, y = total_import)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") 
```

#5) make this more professional! Add labels and a title 

```{r}
ggplot(global_cocoa_import, aes(x = Year, y = total_import)) +
  geom_line(color = "blue") +
  geom_point(color = "blue")  + 
  labs(y= "Total Import (ton)", x= "Year", title = "Global Import of Cocoa from 2015-2019")
```
  

```


