---
title: "Topic 4 activity student version"
output: html_notebook
---
1) #We have 20 office supplies in a box. 10 staplers, 5 paperclips, 2 pens, 3 pencils. Create a vector that shows how many of each office supply is in the box

```{r} 
supplies <- rep(c("staplers", "paperclips", "pens", "pencils"), time=c(10, 5, 2, 3)) 
```

2) show the frequency of each office supply in the box with one function

```{r}
table(supplies)
```
3) What are the proportions of each office supply?

```{r}
prop.table(table(supplies))
```
4) Pull a random sample of 4 office supplies from the box. Set seed to 100. What office supplies are pulled from the box?
```{r}
set.seed(100)
sample(supplies, 4)
```
5) a)Pull another sample of 20 office supplies but replace the office supplies each time. Replicate this event 50 times. Create an object for that sample. set seed to 100. 

```{r}
set.seed(100)
more_supplies <- replicate(50, sample(supplies, 20, replace=T))
more_supplies
```

6) Create a table that shows the probability of pulling each supply after 50 trials. Is it similar to our original table? 

```{r}
prop.table(table(more_supplies))
prop.table(table(supplies))
```

7) We have the following temperature data for 10 days in a given month


```{r}
temp <- c(80, 90, 71, 31, 67, 82, 74, 93, 99, 91)
```

8) Based on this data, and assuming a normal distribution, what is the probability that there's one day that is 73 degrees or lower?

```{r}
avg_temp <- mean(temp)
std_temp <- sd(temp)
pnorm(73, avg_temp, std_temp)
```
9) What is the probability that a day falls between 73 and 83 degrees?
```{r}
low_temp <- pnorm(73, avg_temp, std_temp)
high_temp <- pnorm(83, avg_temp, std_temp)

high_temp - low_temp
```
10) What is the probability that there is a day hotter than 83 degrees

```{r}
1 - pnorm(83, avg_temp, std_temp)
```

