---
title: "Topic 3" Plotting with ggplot2 In Class Activity"
output: html_notebook
---

```{r}
library(dplyr)
library(tidyr)
library (ggplot2)
setwd("/Users/aidenstanton/projects/R")
#change the working directory to where you store your data
#file name is Global import of cocoa beans
```

#1) Create a data frame for the Global Import of Cocoa Beans

```{r}
Cocoa <- read.csv("Global import of cocoa beans.csv")
```



#2) summarise the data to show the total import in tonnes per year for all regions
#hint--refer to tutorial 2: group by and summarise

```{r}
total_year <- Cocoa %>% group_by(Year) %>% summarize(total_import = sum(Import.in.tonnes))
```



#3)create the base of a plot graph in ggplot with just the aesthetic layer for x and y

```{r}
ggplot(total_year, aes(x = Year, y = total_import))
```


#4) create a full point graph in ggplot with blue points
```{r}
ggplot(total_year, aes(x = Year, y = total_import)) + geom_point(color = "blue")
```



#5)add a line to your point graph to a line graph
```{r}
ggplot(total_year, aes(x = Year, y = total_import)) + geom_point(color = "blue") + geom_line(color = "blue") 

```



#6) make this more professional! Add labels and a title 

```{r}
ggplot(total_year, aes(x = Year, y = total_import)) + geom_point(color = "blue") + geom_line(color = "blue") + labs(x = "Year", y = "Imports per year", title = "Cocoa imports per year") + theme_dark()
```

  

```


